---
title: "Standardizing burstiness for occurrence count"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standardizing burstiness for occurrence count}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(distributionalDensity)
set.seed(42)
```

## The problem

Burstiness measures how clustered a category's occurrences are within a
text, using the interevent times between successive occurrences.
`dd_burstiness()` uses the finite-size-corrected estimator of Kim & Jo
(2016) by default (`method = "kj"`), which corrects a known bias in the
older Goh & Barabasi (2008) statistic: under random placement, the
uncorrected statistic runs systematically negative when a category has
few occurrences, so it partly restates occurrence count rather than
measuring clustering independent of it.

The Kim & Jo correction removes most of that *mean* bias, but not the
*sampling variance*: burstiness estimates from few occurrences are still
much noisier than estimates from many. Because occurrence count (`k`)
scales with text length at a fixed category rate, this shows up in
practice as an apparent relationship between burstiness and text length
-- even though the underlying formula is exactly invariant to linear
rescaling of positions and has no text-length term in it.

We can see this directly by simulating random (Poisson-null) placements
and comparing burstiness's spread as occurrence count varies, holding
text length fixed:

```{r sim-variance}
n_sim <- 2000
n_words <- 2000

sim_once <- function(k, n_words) {
  pos <- sort(sample.int(n_words, k))
  dd_burstiness(pos, method = "kj")
}

results <- do.call(rbind, lapply(c(3, 5, 10, 20, 50, 100), function(k) {
  vals <- replicate(n_sim, sim_once(k, n_words))
  data.frame(k = k, mean_B = mean(vals), sd_B = sd(vals))
}))
results
```

The spread collapses from `sd` well over 0.5 at `k = 3` occurrences down
to about 0.05 at `k = 100`, holding text length constant. In a corpus
where category rate is roughly stable but text length varies, shorter
texts produce fewer occurrences and therefore noisier, more extreme
burstiness estimates -- a text-length confound in everything but name.

## Two ways to standardize -- and why they're not interchangeable

`standardize` (on `dd_burstiness()`, `distributional_density()`, and
`dd()`) compares the observed burstiness value to a simulated null:
`n_sim` random placements of the same number of occurrences among the
same number of token slots, using the same estimator -- the same
finite-size null the Kim & Jo correction is already built on. There are
two ways to make that comparison, and they answer different questions:

- **`standardize = "z"`** subtracts the null's mean *and* divides by its
  standard deviation. This is an inferential statistic: it asks "how
  surprising is this value if occurrences were placed at random?"
- **`standardize = "center"`** subtracts only the null's mean, without
  dividing by its spread. This is an effect size: it asks "how much more
  (or less) clustered is this than random placement would produce?"

It's tempting to treat `"z"` as simply a length-adjusted version of
burstiness suitable for any cross-text comparison. It isn't -- because
the null's standard deviation itself shrinks as occurrence count grows
(that's exactly the pattern in the table above), dividing by it
reintroduces an occurrence-count dependence of its own. We can see this
by simulating a *genuinely* bursty process -- not random placement, but
real clustering with a fixed strength -- and varying only how many
occurrences are drawn from it:

```{r sim-bursty-confound}
gen_bursty <- function(k, shape = 0.4, scale = 10) {
  # Weibull(shape < 1) interevent times: genuinely clustered, and the
  # degree of clustering (set by `shape`) does not depend on k.
  iet <- rweibull(k - 1, shape = shape, scale = scale)
  pos <- cumsum(c(0, iet)) + 1
  list(positions = pos, n_words = ceiling(max(pos)) + 5)
}

bursty_results <- do.call(rbind, lapply(c(5, 10, 20, 40, 80), function(k) {
  reps <- replicate(150, {
    g <- gen_bursty(k)
    b <- dd_burstiness(g$positions, method = "kj")
    z <- dd_burstiness(g$positions, standardize = "z", n_words = g$n_words, n_sim = 500)
    ctr <- dd_burstiness(g$positions, standardize = "center", n_words = g$n_words, n_sim = 500)
    c(raw = b, z = z, center = ctr)
  })
  data.frame(k = k, mean_raw = mean(reps["raw", ]),
             mean_z = mean(reps["z", ], na.rm = TRUE),
             mean_center = mean(reps["center", ], na.rm = TRUE))
}))
bursty_results
```

`mean_raw` and `mean_center` stay roughly flat across `k` -- the same
underlying clustering strength reads as the same burstiness regardless
of how many occurrences happened to be observed. `mean_z` climbs
steadily, roughly doubling every time `k` doubles, for a process whose
true clustering never changed. **The z-score reflects the strength of
evidence against random placement, not the size of the clustering
effect** -- more occurrences make randomness easier to rule out even
when the effect itself is constant, the same way a t-statistic grows
with sample size at a fixed effect size. Using `"z"` to compare
clustering magnitudes across texts of different occurrence counts would
reintroduce exactly the confound `standardize` exists to remove.

## Using `standardize = "center"`

```{r standardize-example}
text <- "I lift my mug and I breathe in my first sip of it, mine alone, made for me."
self_words <- c("i", "me", "my", "mine", "myself")

dd(text = text, dictionary = self_words)
dd(text = text, dictionary = self_words, standardize = "center", n_sim = 2000)
```

`burstiness` changes from the raw `[-1, 1]` Kim & Jo value to a
mean-centered value against the simulated null (unbounded, but on
roughly the same scale as the raw statistic); `burstiness_raw` -- the
uncorrected Goh & Barabasi statistic, kept for comparison -- is never
standardized.

## When to use which

- Use `standardize = "center"` when burstiness will be **compared
  across texts** that differ in length or occurrence count -- e.g.
  correlating burstiness with other variables, or intercorrelating it
  with prevalence, position, and dispersion, where a residual
  occurrence-count dependence would otherwise masquerade as a
  substantive relationship. This is the option that behaves like an
  effect size.
- Use `standardize = "z"` only when the question is genuinely
  inferential -- e.g. "is this text's burstiness distinguishable from
  random placement at all?" -- and not when comparing magnitudes across
  texts with different occurrence counts.
- Use the default `standardize = "none"` when you want the literature's
  published Kim & Jo statistic itself -- e.g. reporting burstiness
  values that should read on the same `[-1, 1]` scale used elsewhere in
  the burstiness literature, or reproducing another study's numbers.
- `standardize = "z"`/`"center"` are stochastic (they vary slightly run
  to run unless you `set.seed()` first) and more expensive to compute,
  since they simulate `n_sim` null placements per text-by-category
  combination. Increase `n_sim` for more stable estimates at the cost of
  runtime; 1000 (the default) is adequate for most exploratory use, but
  consider a larger value (e.g. 5000-10000) for a final analysis.
- Burstiness -- standardized or not -- is only defined for categories
  with at least 3 occurrences in a text (`NA` otherwise), and
  `standardize = "z"` additionally returns `NA` if the simulated null
  has zero variance (e.g. occurrences fill nearly every token slot).

## References

Goh, K.-I., & Barabasi, A.-L. (2008). Burstiness and memory in complex
systems. *Europhysics Letters*, *81*(4), 48002.
<https://doi.org/10.1209/0295-5075/81/48002>

Kim, E.-K., & Jo, H.-H. (2016). Measuring burstiness for finite event
sequences. *Physical Review E*, *94*(3), 032311.
<https://doi.org/10.1103/PhysRevE.94.032311>
