---
title: "Standardizing burstiness for occurrence count"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standardizing burstiness for occurrence count}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(distributionalDensity)
set.seed(42)
```

## The problem

Burstiness measures how clustered a category's occurrences are within a
text, using the interevent times between successive occurrences.
`dd_burstiness()` uses the finite-size-corrected estimator of Kim & Jo
(2016) by default (`method = "kj"`), which corrects a known bias in the
older Goh & Barabasi (2008) statistic: under random placement, the
uncorrected statistic runs systematically negative when a category has
few occurrences, so it partly restates occurrence count rather than
measuring clustering independent of it.

The Kim & Jo correction removes most of that *mean* bias, but not the
*sampling variance*: burstiness estimates from few occurrences are still
much noisier than estimates from many. Because occurrence count (`k`)
scales with text length at a fixed category rate, this shows up in
practice as an apparent relationship between burstiness and text length
-- even though the underlying formula is exactly invariant to linear
rescaling of positions and has no text-length term in it.

We can see this directly by simulating random (Poisson-null) placements
and comparing burstiness's spread as occurrence count varies, holding
text length fixed:

```{r sim-variance}
n_sim <- 2000
n_words <- 2000

sim_once <- function(k, n_words) {
  pos <- sort(sample.int(n_words, k))
  dd_burstiness(pos, method = "kj")
}

results <- do.call(rbind, lapply(c(3, 5, 10, 20, 50, 100), function(k) {
  vals <- replicate(n_sim, sim_once(k, n_words))
  data.frame(k = k, mean_B = mean(vals), sd_B = sd(vals))
}))
results
```

The spread collapses from `sd` well over 0.5 at `k = 3` occurrences down
to about 0.05 at `k = 100`, holding text length constant. In a corpus
where category rate is roughly stable but text length varies, shorter
texts produce fewer occurrences and therefore noisier, more extreme
burstiness estimates -- a text-length confound in everything but name.

## The fix

`standardize = TRUE` (on `dd_burstiness()`, `distributional_density()`,
and `dd()`) replaces the raw Kim & Jo value with a z-score computed
against a simulated null: `n_sim` random placements of the same number
of occurrences among the same number of token slots, using the same
estimator. This is the same finite-size null the Kim & Jo correction is
already built on -- standardizing just finishes correcting for it,
variance and all, rather than only the mean.

```{r standardize-example}
text <- "I lift my mug and I breathe in my first sip of it, mine alone, made for me."
self_words <- c("i", "me", "my", "mine", "myself")

dd(text = text, dictionary = self_words)
dd(text = text, dictionary = self_words, standardize = TRUE, n_sim = 2000)
```

`burstiness` changes from the raw `[-1, 1]` Kim & Jo value to a z-score
(unbounded, roughly standard-normal in scale under the null);
`burstiness_raw` -- the uncorrected Goh & Barabasi statistic, kept for
comparison -- is never standardized.

## Standardized burstiness is no longer confounded with occurrence count

Repeating the simulation above with `standardize = TRUE` shows the mean
stays near zero and, more importantly, the variance no longer depends
on `k`:

```{r sim-standardized}
sim_once_std <- function(k, n_words) {
  pos <- sort(sample.int(n_words, k))
  dd_burstiness(pos, standardize = TRUE, n_words = n_words, n_sim = 300)
}

results_std <- do.call(rbind, lapply(c(3, 5, 10, 20, 50), function(k) {
  vals <- replicate(300, sim_once_std(k, n_words))
  data.frame(k = k, mean_z = mean(vals, na.rm = TRUE), sd_z = sd(vals, na.rm = TRUE))
}))
results_std
```

## When to use which

- Use `standardize = TRUE` when burstiness will be **compared across
  texts** that differ in length or category rate -- e.g. correlating
  burstiness with other variables, or intercorrelating it with
  prevalence, position, and dispersion, where a residual occurrence-count
  dependence would otherwise masquerade as a substantive relationship.
- Use the default `standardize = FALSE` when you want the literature's
  published Kim & Jo statistic itself -- e.g. reporting burstiness
  values that should read on the same `[-1, 1]` scale used elsewhere in
  the burstiness literature, or reproducing another study's numbers.
- `standardize = TRUE` is stochastic (it varies slightly run to run
  unless you `set.seed()` first) and more expensive to compute, since it
  simulates `n_sim` null placements per text-by-category combination.
  Increase `n_sim` for more stable estimates at the cost of runtime;
  1000 (the default) is adequate for most exploratory use, but consider
  a larger value (e.g. 5000-10000) for a final analysis.
- Burstiness -- standardized or not -- is only defined for categories
  with at least 3 occurrences in a text (`NA` otherwise), and
  `standardize = TRUE` additionally requires enough spread in occurrence
  positions for the simulated null to have nonzero variance.

## References

Goh, K.-I., & Barabasi, A.-L. (2008). Burstiness and memory in complex
systems. *Europhysics Letters*, *81*(4), 48002.
<https://doi.org/10.1209/0295-5075/81/48002>

Kim, E.-K., & Jo, H.-H. (2016). Measuring burstiness for finite event
sequences. *Physical Review E*, *94*(3), 032311.
<https://doi.org/10.1103/PhysRevE.94.032311>
